The state learner \\ {-- a super learner for right-censored data}

### 2024-04-29 tex/jrss-version/state-learner-jrss-style.tex
response to TAG's comments

Medical decision making is traditionally based on the discrepancy
between intuitive jugdment and the law of chance
\citep{redelmeier1995probability}. We are concerned with the
statistical models that provide probabilistic predictions about the
medical future of a patient. We have two different applications of
these models in mind. One is as a stand-alone tool where the model
provides information for a freshly diagnosed patient. The other is as
input to an algorithm which estimates a causal target parameter such
as an average treatment effect. The models are learned on data from
previous patients. A super learner is a machine learning algorithm
that combines a finite set of learners into a meta learner by
estimating prediction performance in hold-out samples using a
pre-specified loss function \citep{van2007super}. When the aim is to
make a prediction model, super learners combine strong learners, such
as Cox regression models and random survival forests
\citep{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.

While the general idea of combining strong learners based on
cross-validation data stems from earlier work
\citep{wolpert1992stacked,breiman1996stacked}, the name super learner
is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed
to perform almost as well as the model which minimises the expected
performance, i.e., the model we would select if we could evaluate the
prediction performance in an infinite hold-out sample. The choice of
the loss function for super learning in survival analysis deserves a
closer look. Existing super learner algorithms for right-censored data
use partial log-likelihood loss or inverse probability of censoring
weighted loss
\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
The use of the partial log-likelihood loss, however, restricts the
class of learners and excludes for example simple Kaplan-Meier based
learners and also more complex random survival forest algorithms. For
this reason \cite{golmakani2020super} restrict their learners to Cox
proportional hazard models. A lesser known fact is that a super
learner constructed with the negative partial log-likelihood loss
implicitly depends on the censoring distribution
\citep{hjort1992inference,whitney2019comment}. Inverse probability of
censoring weighted loss has been studied for decades
\citep{graf1999assessment, hothorn2006survival, gerds2006consistent}. A disadvantage of inverse probability of
censoring weighted loss estimators is that they require a
pre-specified model for the censoring
distribution. In \cite{westling2021inference} this problem is tackled by
iterating between super learning of the censoring distribution and the
event time
distribution. % ANDERS: mangler afrundning efter sidste saetning. eller maaske kan saetningen flyttes laenger ned?

In this article we define the state learner, a new super learner for
right-censored data, which simultaneously estimates the expected loss
of learners of the event time distribution and the censoring
distribution. The loss function which is used to define the state
learner is based on the observable
quantities. % ANDERS: hvad mener vi med observable quantities? distribution of the observed random variables?
The state learner can be applied to all types of survival models,
works in the presence of competing risks, and does not require a
single pre-specified estimator of the conditional censoring
distribution.  To analyze the theoretical properties of the state
learner we focus on the discrete super learner which combines the
library of learners by picking the one that minimises the
cross-validated loss
\citep{van2011targeted}. % ANDERS: er der bedre referencer fra Mark paper/bog for discrete super learner?
In the presence of competing risks, our algorithm uses separate
libraries to model the hazard rates of the competing risks and the
censoring distribution. We show that the oracle selector of the state
learner is consistent if all libraries contain a consistent learner
and prove a finite sample oracle inequality. The result of the state
learner algorithm is a model which predicts the probability of an
event based on covariates in the presence of competing risks
\citep{gerds2021medical}. The other application is in targeted
learning where conditional event probabilities occur as
high-dimensional nuisance parameters which need to be estimated at a
certain rate \citep{van2011targeted, rytgaard2021estimation,
  rytgaard2022targeted}. We show for a targeted estimator, which uses
the state learner to estimate nuisance parameters, that a second order
product structure holds for the asymptotic bias term.

@@

In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
time to event variable, \(D\in\{1,2\}\) the cause of the event, and
$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
continuous for all \( x \) and \( j \). %ANDERS: need a note here which says that this is not a limitation?
We denote by \(S\) the conditional
event-free survival function:
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
% ANDERS: I prefer to have the \tau explicit
Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
measures on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
that the event times and the censoring times are conditionally
independent given covariates, \( T \independent C \mid X \). This
implies that any distribution \( P \in \mathcal{P} \) is characterised
by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
hazard function for \( C \) given \( X \)
\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
function for censoring. % ANDERS: can we not allow Gamma to be discrete?
For notational convenience, we assume that \( \Gamma(\blank \mid x) \) is
continuous for all \( x \), and let
\(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. The distribution 
is characterized by
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
% ANDERS: H is missing in the model?
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
also have
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

@@

Machine learning based on right-censored data commonly use the
negative partial log-likelihood as loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
It seems however that this loss function is not perfectly suited for
super learning, because it cannot be applied to important survival
learners: The Kaplan-Meier estimator, the random survival forest, and
the semi-parametric Cox regression model provide cumulative hazard
functions that are piece-wise constant in the time argument, and hence
assign zero probability to event times not observed in the training
data. This implies that when event times are observed in continuous
time, any of these learners will almost surely have infinite loss in
an independent test set according to the negative partial
log-likelihood loss.
% ANDERS: would be good to have a formula for a the negative partial log-likelihood loss showing that the result is infitity

When a proportional hazards model is assumed, the baseline hazard
function can be profiled out of the likelihood to give a new partial
log-likelihood loss \citep{cox1972regression}, which has been
suggested as a loss function for super learning
\citep{golmakani2020super,verweij1993cross}. While this allows the
library of learners to include Cox' proportional hazard models, the
drawback is that the library is now restricted to include only these
models. % ANDERS: can the library really include multiple proportional hazards models? only one of them can be correct ...
% also, would be good to merge this paragraph with the previous somehow

@@

\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations
in equation~(\ref{eq:parametrizeP}), to learn a feature of the
observed data distribution \( P \). The discrete state learner ranks a
tuple of learners of \( (\Lambda_1, \Lambda_2, \Gamma) \) based on how
well they jointly model the observed data. %ANDERS: would be good to add one sentence about how to obtain a prediction model from such a ranking.
To formally introduce the
state learner, we define the multi-state process

@@

% ANDERS: Figuren er lavet med prodlim, ik? du kan angive nummer for boksene (0 = initial, 1 = cause 1, 2= cause 2 og jeg ville foretraekke 3 i stedet for -1 for censored)

@@

% ANDERS: the next sentence requires conditional independent censoring and restriction to [0,\tau], right?
Because each quadruple

@@

\(P\in\mathcal P\) which in turn determines \( (F, H) \),

@@

a learner for \( F \) can be constructed from learners of \( \Lambda_1 \),
\( \Lambda_2 \), and $\Gamma$ as follows:

@@

In this section we establish theoretical guarantees for the state learner.
% We let \( \E_P \) denote
% expectation under \( P \) and use \( \Theta \) to denote the
% collection of all conditional state-occupation probability functions.
Proposition~\ref{prop:stric-prop} can be derived from the fact that
the integrated Brier score (also called the continuous ranked
probability score) is a strictly proper scoring rule
\citep{gneiting2007strictly}. This implies that if we minimise the
average loss of the integrated Brier score, we recover the parameters
of the data-generating distribution. Specifically, this implies that
the oracle of a state learner is consistent if the library of learners
contains at least one learner that is consistent for estimation of
\( F \). Recall that the function \(F\) implicitly depends on the data
generating probability measure \(P\in\mathcal P\) but that this was
suppressed in the notation. We now make this dependence explicit by
writing \(F_0\) for the function which is obtained by substituting a
specific \(P_0\in\mathcal{P}\) for \(P\) in equation
\eqref{eq:transition}.

\begin{proposition}
  \label{prop:stric-prop}
  For \(P_0\in\mathcal{P}\) define
  \begin{equation*}
    F^* = \argmin_{F} P_0{[\bar{B}_\tau(F, \blank)]},
  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
  \end{equation*}
  %ANDERS: would be good to define the set over which F is minimized with a symbol rather than only words
where the minimum is taken over all \( F \), such that \( F \) is a conditional
state occupation probability function for some measure \( P \) as defined in equation~(\ref{eq:F-def}).
Then \( F^*(t,j,\blank) = F_0(t,j,\blank) \) \( H \)-almost surely for
any \( j\in \{-1,0,1,2\} \) and almost any \( t \in [0, \tau]\).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

We now establish a finite sample oracle result for the state learner.
Our Corollary~\ref{cor:oracle-prop} is in essence a special case of Theorem 2.3
in \cite{vaart2006oracle}. % ANDERS: is Theo 2.3 correct?
We assume
that we split the data into equally sized folds, and for simplicity of
presentation we take \( n \) to be such that
\( |\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow the
number of learners to grow with \( n \) and write
\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
\mathcal{B}_n)\) as short-hand notation and to emphasise the
dependence on \( n \).
% The discrete super $\hat{\phi}_n$ and the
% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
% but now with respect to the library \( \mathcal{F}_n \) and the loss
% \( B \).
In the following we let \( \| \blank \|_{P_0} \) denote the norm
% ANDERS: norm on which space? 
\begin{equation}
  \label{eq:norm}
  \| F \|_{P_0} = 
  \left\{
    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
  \right\}^{1/2}.
\end{equation}
\begin{corollary}
  \label{cor:oracle-prop}
  For all \(P_0\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
  and $\delta>0$,
  \begin{align*}
    \E_{P_0}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
    & \leq (1 + 2\delta)
      \E_{P_0}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_0 \Vert_{P_0}^2 \right]}
    \\
    & \quad
      + (1+ \delta) 16   K \tau
      \left(
      13 + \frac{12}{\delta}
      \right)
      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
  \end{align*}
\end{corollary}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

>>>>

Medical decision making is traditionally based on the discrepancy between
intuitive judgment and the law of chance \citep{redelmeier1995probability}. We
are concerned with the statistical models that provide probabilistic predictions
about the medical future of a patient. We have two different applications of
these models in mind.
% THOMAS: I think we only really have one in mind in this article. For
% stand-alone prediction, I think some further things should be
% addressed/discussed, which I had thought could be for another article. But
% maybe we can but it in here.
One is as a stand-alone tool where the model provides
information for a freshly diagnosed patient. The other is as input to an
algorithm which estimates a causal target parameter such as an average treatment
effect. The models are learned on data from previous patients. A super learner
is a machine learning algorithm that combines a finite set of learners into a
meta learner by estimating prediction performance in hold-out samples using a
pre-specified loss function \citep{van2007super}. When the aim is to make a
prediction model, super learners combine strong learners, such as Cox regression
models and random survival forests \citep{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.

While the general idea of combining strong learners based on
cross-validation data stems from earlier work
\citep{wolpert1992stacked,breiman1996stacked}, the name super learner
is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed
to perform almost as well as the model which minimises the expected
performance, i.e., the model we would select if we could evaluate the
prediction performance in an infinite hold-out sample. The choice of
the loss function for super learning in survival analysis deserves a
closer look. Existing super learner algorithms for right-censored data
use partial log-likelihood loss or inverse probability of censoring
weighted loss
\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
The use of the partial log-likelihood loss, however, restricts the
class of learners and excludes for example simple Kaplan-Meier based
learners and also more complex random survival forest algorithms. For
this reason \cite{golmakani2020super} restrict their learners to Cox
proportional hazard models. A lesser known fact is that a super
learner constructed with the negative partial log-likelihood loss
implicitly depends on the censoring distribution
\citep{hjort1992inference,whitney2019comment}. Inverse probability of
censoring weighted loss has been studied for decades
\citep{graf1999assessment, hothorn2006survival, gerds2006consistent}. A disadvantage of inverse probability of
censoring weighted loss estimators is that they require a
pre-specified model for the censoring
distribution. In \cite{westling2021inference} this problem is tackled by
iterating between super learning of the censoring distribution and the
event time
distribution. 
% ANDERS: mangler afrundning efter sidste saetning. eller maaske kan saetningen
% flyttes laenger ned?
% 
% THOMAS: Tror du har skrevet den her bid. Har forsøgt at afrunde.
It is not clear, however, whether such a procedure converges towards an element
minimizing the average loss.

In this article we define the state learner, a new super learner for
right-censored data, which simultaneously estimates the expected loss
of learners of the event time distribution and the censoring
distribution. The loss function which is used to define the state
learner is based on the observable
variables. % ANDERS: hvad mener vi med observable quantities? distribution of
% the observed random variables?
% 
% THOMAS: Changed to 'variables'. It is important that the loss is based on the
% observed variables directly, not just their distribution.
The state learner can be applied to all types of survival models,
works in the presence of competing risks, and does not require a
single pre-specified estimator of the conditional censoring
distribution.  To analyze the theoretical properties of the state
learner we focus on the discrete super learner which combines the
library of learners by picking the one that minimises the
cross-validated loss
\citep{van2007super}.
% ANDERS: er der bedre referencer fra Mark paper/bog for discrete super learner?
% 
% THOMAS: Yes, har taget Hubbard+laan super learner article i stedet.
In the presence of competing risks, our algorithm uses separate
libraries to model the cumulative hazard functions of the competing risks and the
censoring distribution. We show that the oracle selector of the state
learner is consistent if all libraries contain a consistent learner
and prove a finite sample oracle inequality. The result of the state
learner algorithm is a model which predicts the probability of an
event based on covariates in the presence of competing risks
\citep{gerds2021medical}.
% THOMAS: This is not the output directly, but can easily be obtained.
The other application is in targeted
learning where conditional event probabilities occur as
high-dimensional nuisance parameters which need to be estimated at a
certain rate \citep{van2011targeted, rytgaard2021estimation,
  rytgaard2022targeted}. We show for a targeted estimator, which uses
the state learner to estimate nuisance parameters, that a second order
product structure holds for the asymptotic bias term.

@@

In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
time to event variable, \(D\in\{1,2\}\) the cause of the event, and
$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
continuous for all \( x \) and \( j \). 
% ANDERS: need a note here which says that this is not a limitation?
% 
% THOMAS: I think it is already there in 'For ease of notation...', but I have
% added the next sentence. We could also just work with the product integral,
% would that be better?
This is not a limitation as everything caries over directly to the general case
and is solely imposed to simplify the formulas. We denote by \(S\) the
conditional event-free survival function:
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
% ANDERS: I prefer to have the \tau explicit
% 
% THOMAS: OK
Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
measures on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
that the event times and the censoring times are conditionally
independent given covariates, \( T \independent C \mid X \). This
implies that any distribution \( P \in \mathcal{P} \) is characterised
by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
hazard function for \( C \) given \( X \)
\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
function for censoring.
% ANDERS: can we not allow Gamma to be discrete?
% 
% THOMAS: Yes, no problem, it is only for notation convenience. 
For notational convenience, we assume that \( \Gamma(\blank \mid x) \) is
continuous for all \( x \), and let
\(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. The distribution 
is characterized by
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
% ANDERS: H is missing in the model?
% 
% THOMAS: H is part of Q (see equation between equation 1 and 2).
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
also have
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

@@

Machine learning based on right-censored data commonly use the
negative partial log-likelihood as loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
It seems however that this loss function is not perfectly suited for
super learning, because it cannot be applied to important survival
learners: The Kaplan-Meier estimator, the random survival forest, and
the semi-parametric Cox regression model provide cumulative hazard
functions that are piece-wise constant in the time argument, and hence
assign zero probability to event times not observed in the training
data. This implies that when event times are observed in continuous
time, any of these learners will almost surely have infinite loss in
an independent test set according to the negative partial
log-likelihood loss.
% ANDERS: would be good to have a formula for a the negative partial
% log-likelihood loss showing that the result is infitity
%
% THOMAS: I would argue that this should be clear from the previous sentence and
% the definition of the log-likelihood function: If a discrete model assigns a
% probability of zero to an event, then the log-likelihood of that event is
% log(0) = -infinity.
%
% Such a formula was part of one of the earlier versions where I had put more
% focus on the problem than on the solution. If we add this, we should work with
% the product integral notation.

When a proportional hazards model is assumed, the baseline hazard
function can be profiled out of the likelihood to give a new partial
log-likelihood loss \citep{cox1972regression}, which has been
suggested as a loss function for super learning
\citep{golmakani2020super,verweij1993cross}. While this allows the
library of learners to include Cox' proportional hazard models, the
drawback is that the library is now restricted to include only these
models.
% ANDERS: can the library really include multiple proportional hazards models?
% only one of them can be correct ... also, would be good to merge this
% paragraph with the previous somehow
%
% THOMAS: Yes, it can include multiple PH models (if I understand you
% correctly). A major point is that there is no proof that this works, and when
% you start to think about it, it is not completely obvious whether it should
% work or not. I think it works and am trying to prove it.
%
% I disagree about merging this section with the previous one, I see them as
% describing two completely different approaches and issues.

@@

\( \Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
distribution \( P \). The discrete state learner ranks a tuple of learners of
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
observed
data.
% ANDERS: would be good to add one sentence about how to obtain a prediction
% model from such a ranking.
%
% THOMAS: I think this depends on whether or not we want prediction to be one of
% the tasks addressed in this article. I have focused on something else. We
% could add something like:
Risk predictions for causes one and two can then be obtained from
\( \Lambda_1 \) and $\Lambda_2$ \citep{benichou1990estimates}. To formally
introduce the state learner, we define the multi-state process

@@

% ANDERS: Figuren er lavet med prodlim, ik? du kan angive nummer for boksene (0
% = initial, 1 = cause 1, 2= cause 2 og jeg ville foretraekke 3 i stedet for -1
% for censored)
%
% THOMAS: Jeg foretrækker -1 frem for 3, så vil gerne høre dine argumenter for
% 3.

@@

% ANDERS: the next sentence requires conditional independent censoring and
% restriction to [0,\tau], right?
%
% THOMAS: Yes, added now.
Under conditional independent censoring each quadruple

@@

\(P\in\mathcal P\), c.f.\ equation~\eqref{eq:parametrizeP}, which in turn
determines \( (F, H) \),

@@

%
% THOMAS: Yes, added above -- OK?
and so a learner for \( F \) can be constructed from learners of
\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:

@@

In this section we establish theoretical guarantees for the state learner. We
show that the state learner is consistent if its library contains a consistent
learner, and we establish a finite sample inequality for the excess risk of the
state learner compared to the oracle. Finally we show that a certain second
order structure is preserved when the state learner is used for targeted
learning.

\subsection{Consistency}
\label{sec:consistency}

% We let \( \E_P \) denote
% expectation under \( P \) and use \( \Theta \) to denote the
% collection of all conditional state-occupation probability functions.
Proposition~\ref{prop:stric-prop} can be derived from the fact that the
integrated Brier score (also called the continuous ranked probability score) is
a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
if we minimise the average loss of the integrated Brier score, we recover the
parameters of the data-generating distribution. Specifically, this implies that
the oracle of a state learner is consistent if the library of learners contains
at least one learner that is consistent for estimation of \( F \). Recall that
the function \(F\) implicitly depends on the data generating probability measure
\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
this dependence explicit by writing \(F_0\) for the function which is obtained
by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
\eqref{eq:transition}. In the following we let
\( \mathcal{H} = \{F_P : P \in \mathcal{P}\} \) where \( F_P \) is defined as in
equation~(\ref{eq:F-def}) using the measure \( P \in \mathcal{P} \).

\begin{proposition}
  \label{prop:stric-prop}
  For \(P_0\in\mathcal{P}\) define
  \begin{equation*}
    F^* = \argmin_{F \in \mathcal{H}} P_0{[\bar{B}_\tau(F, \blank)]}.
  % F^* = \argmin_{(\Lambda_1,\Lambda_2,\Gamma) \in \mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}\otimes\mathcal{M}_{\tau}} P_0{[\bar B_\tau(\phi_{(\Lambda_1,\Lambda_2,\Gamma)}, \blank)]},
  \end{equation*}
  % ANDERS: would be good to define the set over which F is minimized with a
  % symbol rather than only words
  %
  % THOMAS: How about now? \mathcal{H} is now defined just before this proposition.
Then \( F^*(t,j,\blank) = F_0(t,j,\blank) \) \( H \)-almost surely for
any \( j\in \{-1,0,1,2\} \) and almost any \( t \in [0, \tau]\).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:proof-proposition}.
\end{proof}

\subsection{Oracle inequalities}
\label{sec:finite-sample-oracle}

We establish a finite sample oracle result for the state learner. Our
Corollary~\ref{cor:oracle-prop} is in essence a special case of Theorem 2.3 in
\cite{vaart2006oracle}.
% ANDERS: is Theo 2.3 correct?
%
% THOMAS: Yes. What do you mean?
We assume that we split the data into equally sized folds, and for simplicity of
presentation we take \( n \) to be such that \( |\data_n^{-k}| = n/K \) with
\( K \) fixed. We will allow the number of learners to grow with \( n \) and
write
\( \mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
\mathcal{B}_n)\) as short-hand notation and to emphasise the dependence on
\( n \).
% The discrete super $\hat{\phi}_n$ and the
% oracle $\tilde{\phi}_n$ are defined as in Section~\ref{sec:framework}
% but now with respect to the library \( \mathcal{F}_n \) and the loss
% \( B \).
In the following we let the space \( \mathcal{H} \) be equipped with the norm
\( \| \blank \|_{P_0} \) defined as
% ANDERS: norm on which space?
%
% THOMAS: Fixed.
\begin{equation}
  \label{eq:norm}
  \| F \|_{P_0} = 
  \left\{
    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t H_0( \diff x)
  \right\}^{1/2}.
\end{equation}
